Ethical Considerations of Artificial Intelligence via Neural Networks Applied to Medical Applications

### 2020-05-05 paper.tex
A whole bunch of stuff

	\begin{abstract}
		Placeholder abstract
	\end{abstract}

@@

		Artificial intelligence (AI) and, more particularly, machine learning (ML) is becoming increasingly prevalent in medical fields. Though regulations are being discussed regarding AI as a whole, the regulations proposed by the FDA\cite{fdaregproposal} place a large emphasis on procedural AI which leaves the capabilities of ML models relatively less examined. Simply refusing to implement those restrictions would cause more harm than good, however. We wish to add to this discussion in the hopes of extending the ethical reasoning and guidance with regards to neural network implementations, so as to achieve more safety in applications of these devices and to greater understand the realm of applicability of these devices.

		% Change gets overused in this paragraph.
		Machine learning comprises a subset of AI that closely matches what most commonly comes to mind when AI is mentioned---both conversationally and within public culture.\textbf{[Citation needed]} The difference between procedural AI and ML is detailed in section \ref{introtoai}, but, in essence, procedural AI takes input data and outputs a response by following a strict set of rules or procedures, while ML takes input data and outputs a response determined by a set of facets of interest and weights that may change and shift in response to each input. Frequently ML models undergo a period of "training", during which weights may shift wildly, with the rate of change slowing down until it reaches a period of relatively little change.\textbf{[Citation needed]}

		We also provide an overview of ethical considerations within the context of medical devices in section \ref{ethics} before continuing to weigh the pros and cons of machine learning. We take survey of a number of nursing students to gauge their general opinion, understanding, and expectations of ML. We believe, due to how overworked nurses can be\cite{doi:10.1111/j.1365-2648.2009.05082.x} and how much of an impact ML has had and will have on the nursing population\cite{Ham2017}, that their viewpoints are and important part of the conversation which we wish to highlight. Finally, in section \ref{market}, we delve into the role both regulation and the free market can play in this developing field, and attempt to strike a balance between the two that we hope could create a safe environment for patients and practitioners alike without stifling the development of more refined machine learning and artificial intelligence.

		\subsection{Review of Terminology}
			% This part is definitely really rough, and needs to be expanded. It should potentially be moved to other situations. Probably to the intro to AI section.
			First however, we must define some terms used commonly in the ML field that may be misleading to those without a background in it. There are a number of terms used throughout the field of AI that may be misleading at first glance: "feature" refers to what amounts to an interesting property or value in the input data represented by a function explicitly defined by the developers of the AI. Within the context of ML, these functions are what is modified by changes in the weights of the model.\textbf{[Citation needed?]} Additionally, the term area under the curve (AUC) is used to express the quality of predictions derived from values referred to as sensitivity, the rate of positive predictions (true or false), and selectivity, the rate of negative predictions (true or false). It is important to note that the terms positive and negative apply to predictions, while true and false apply exclusively to the reality of the situation.\cite{introtoauc}

			Explainability of ML models is an important ongoing research field. An explanation of a model seeks to convey why a model made the decisions it did and why it outputted what it did, and explanations have two key attributes: completeness (how accurately the explanation conveys the workings of the ML model), and interpretability (how easy it is for humans to understand the explanation). The two are often at odds, and finding the solution to this problem is an important field of study\cite{8631448}---especially regarding the validation of ML model outputs.\cite{10.1145/3328519.3329126} Unfortunately, the term interpretability is not only applied to explanations, but also to the ML models themselves. When used to describe a model, the term refers to how easy it is to identify and understand the mechanisms that drive the machine. Essentially interpretability seeks to answer how the machine works, while explainability regards why the machine did what it did.\cite{8631448} More in depth definitions of these concepts are given in section \ref{introtoai}.

			% This paragraph should probably be moved.
			This topic also leads to a number of ethical considerations (further discussed in section \ref{ethics}) as pointed out by Gilpin \emph{et al.} who state that "[they] believe that it is fundamentally unethical to present a simplified description of a complex system in order to increase trust if the limitations of the simplified description cannot be understood by users". They point out that an explanation that meets that criteria are both misleading and dangerous.\cite{8631448} Though they do not mention it, we would like to add that though it may often be unintentional, this could easily be used to intentionally deceive.

@@

		When discussing the potential benefits of machine learning for use in medical settings, it's easy to assume that use cases will be centered around aiding health care professionals in decision making and diagnosis or patients in seeking the correct medical care, but there is just as much potential for machine learning to alleviate some of the stress of day to day nursing work simply by streamlining existing systems in ways that match any given situation. In 2016, Ham \emph{et al.} proposed and developed a proof of concept application that provides contextual assistance using wi-fi based location data to aid a machine learning algorithm.\cite{Ham2017}

	\section{Risks of Medical AI}
		Within public media, perception of the risk of ML is often centered entirely around the threat of some kind of AI uprising.\cite{bbc2016rroai} In contrast, the largest threats that concern professionals are, shown by a survey conducted by Danial Faggella in 2019, the lack of explainability and verifiability in most machine learning models complicated enough to be of use in medicine, along with (though not strictly related to medicine) the potential socio-economic and environmental impact of increased use of machine learning.\cite{emerj2019roawrtiwwa}

@@

		\subsection{Lack of Explainability in Sufficiently Complex Systems}
			A key issue lies in two important facets of machine learning: explainability of machine learning models and interpretability of their results. Luo \emph{et al.} define explainability as the ability to "summarize the reasons for the behavior of [machine learning] algorithms" and interpretability as the ability to "comprehend what a model did."\cite{doi:10.1259/bjro.20190021} Though the difference between the two is subtle, the distinction is still an important one to make; methods that may increase interpretability may not yield results regarding the overall explainability of the model. However, improvements to explainability increase interpretability by definition.

@@

		\subsection{Difficulty of Validation}
			There is immense difficulty of validating the predictions (before acting on them) of a machine learning model. This creates further risk by obfuscating potential error behind a highly technical veil for non-machine learning experts. Moreover, unknown error introduced in input datasets is difficult to detect without access to the data itself, as the ML model will continue to output seemingly valid predictions\cite{10.1145/3328519.3329126} Error could be introduced through any number of innocuous vectors (differences in storage systems between health centers, errors digitizing physical records, decay of electronic records over time) or through more malignant ones (such as a targeted attack). The ethical delicacy of handling electronic health records\cite{ford2016privacy} can make impossible to directly examine new inputs for a source of error, and even if it were possible, the sheer magnitude of the task would be prohibitive. Redyuk \emph{et al.} proposes a method of detecting dataset shifts (without direct examination) that may mitigate this issue in the future\cite{10.1145/3328519.3329126}, but it still remains something important to discuss.

@@

		\subsection{Diffusion of Liability}
			Though largely considering more concrete applications of machine learning (such as self driving cars), Reed {et al.}'s 2016 article on the legal responsibility and liability (with regards to English law) points out that there are several parties that could be considered at fault should harm come to a patient due to the complex nature of development and usage of machine learning models. Moreover they posit that it would be incredibly difficult to establish the fault of one party and significantly more expensive to establish the fault of several parties in unison.\cite{reed2016responsibility}

@@

@@

		%\subsection{Balance of Regulation Against the Free Market}\label{market}
			% This section will go into detail on the pros and cons of high and low regulation, emphasizing both the massive increase in risk that comes with low regulations and how high regulations could strangle innovation.

>>>>

	\begin{abstract}
		We discuss a number of ethical considerations regarding emerging machine learning technologies in medicine, as well as briefly outlining machine learning for those not acquainted with it. Additionally, we cover some of the benefits of present and potential applications of machine learning in medicine. We intend to call attention to and discuss issues such as explainability in machine learning and its impact both on the field of machine learning and patients. We hope to provide insight that doesn't require a background in machine learning to understand.
	\end{abstract}

@@

		Artificial intelligence (AI) and, more particularly, machine learning is becoming increasingly prevalent in medical fields. Though regulations are being discussed regarding AI as a whole, the regulations proposed by the FDA\cite{fdaregproposal} place a large emphasis on procedural AI which leaves the capabilities of ML models relatively less examined. Simply refusing to implement those restrictions would cause more harm than good, however. We wish to add to this discussion in the hopes of extending the ethical reasoning and guidance with regards to neural network implementations, so as to achieve more safety in applications of these devices and to greater understand the realm of applicability of these devices.

		% Change gets overused in this paragraph.
		Machine learning comprises a subset of AI that closely matches what most commonly comes to mind when AI is mentioned---both conversationally and within public culture. The difference between procedural AI and machine learning is detailed in section \ref{introtoai}, but, in essence, procedural AI takes input data and outputs a response by following a strict set of rules or procedures, while machine learning takes input data and outputs a response determined by a set of facets of interest and weights that may change and shift in response to each input. Frequently ML models undergo a period of "training", during which weights may shift wildly, with the rate of change slowing down until it reaches a period of relatively little change.

		We also provide an overview of ethical considerations within the context of medical devices in section \ref{ethics} before continuing to weigh the pros and cons of machine learning. We initially intended take survey of a number of nursing students to gauge their general opinion, understanding, and expectations of machine learning. We believe, due to how overworked nurses can be\cite{doi:10.1111/j.1365-2648.2009.05082.x} and how much of an impact machine learning has had and will have on the nursing population\cite{Ham2017}, that their viewpoints are and important part of the conversation which we wish to highlight. However, we had to abort this plan due to the ongoing pandemic at the time of writing.

		\subsection{Review of Terminology}
			% This part is definitely really rough, and needs to be expanded. It should potentially be moved to other situations. Probably to the intro to AI section.
			First however, we must define some terms used commonly in the field of machine learning that may be misleading to those without a background in it. The term "feature" refers to what amounts to an interesting property or value in the input data represented by a function explicitly defined by the developers of the AI. Within the context of machine learning, these functions are what is modified by changes in the weights of the model. Additionally, the term area under the curve (AUC) is used to express the quality of predictions derived from values referred to as sensitivity, the rate of positive predictions (true or false), and selectivity (or specificity), the rate of negative predictions (true or false). It is important to note that the terms positive and negative apply to predictions, while true and false apply exclusively to the reality of the situation. As an example, a test with high sensitivity identify a large number of known positives, while a test with high selectivity would negatively identify a large number of known negatives.\cite{introtoauc,econ2020sense}

			Explainability of machine learning models is an important ongoing research field. An explanation of a model seeks to convey why a model made the decisions it did and why it outputted what it did, and explanations have two key attributes: completeness (how accurately the explanation conveys the workings of the ML model), and interpretability (how easy it is for humans to understand the explanation). The two are often at odds, and finding the solution to this problem is an important field of study\cite{8631448}---especially regarding the validation of ML model outputs.\cite{10.1145/3328519.3329126} Unfortunately, the term interpretability is not only applied to explanations, but also to the machine learning models themselves. When used to describe a model, the term refers to how easy it is to identify and understand the mechanisms that drive the machine. Essentially explainability seeks to answer how the machine works, while interpretability regards why the machine did what it did.\cite{8631448} More in depth definitions of these concepts are given in section \ref{introtoai}.

			% This paragraph should probably be moved.
			% This topic also leads to a number of ethical considerations (further discussed in section \ref{ethics}) as pointed out by Gilpin \emph{et al.} who state that "[they] believe that it is fundamentally unethical to present a simplified description of a complex system in order to increase trust if the limitations of the simplified description cannot be understood by users". They point out that an explanation that meets that criteria are both misleading and dangerous.\cite{8631448} Though they do not mention it, we would like to add that though it may often be unintentional, this could easily be used to intentionally deceive.

@@

		Artificial Intelligence can improve society's overall health. Researchers are mainly focused on using AI for cancer, neurological diseases, and cardiological diseases. Its usages in healthcare can be broken down into three categories: clinical practice, translational research, and biomedical research.

		Currently the focus of AI development  in clinical practice is for diagnostic purposes. One of the most successful areas in diagnostics is automated image-based diagnosis. Medical professionals such as ones in radiology, ophthalmology, dermatology, and pathology rely on image based diagnoses. In radiology, applications of AI such as the detection of lung nodules, the diagnosis of pulmonary tuberculosis and breast mass identification have reached expert level diagnostic accuracy. This also applies to dermatologists, whose job is largely based around inspection of the skin and its irregularities. Typical skin melanoma has visual features that separate it from benign lesions. Dermatologists have even developed a guideline known as ABCDE. A stands for the asymmetry of a leions, B for irregular borders, C for color variegation, D for a diameter of 6 mm or greater, and E for enlargement of the surface of a lesion. With the exception of E, every other guideline can be accessed from a photo. A neural network was trained on 129,450 images to identify skin melanoma. The accuracy of the neural network was greater than the average dermatologists based a comparison between the assessments of the neural network and twenty-one dermatologists. The training for the neural network was computationally intensive, but could be deployed as a mobile app granting in home access to expert level screening. In ophthalmology, retinal cameras are used to capture the retina, optic disc,and macula. This is used to detect and monitor diseases such as DR, glaucoma, neoplasms of the retina and age-related macular degeneration. Artificial intelligence helps when it comes to diabetic patients. Diabetic patients are at risk for DR and are recommended to be screened every year. Usually, an ophthalmologist will examine and interpret the photos. This creates a large workload for ophthalmologists for something that is a part of a routine check up for many people. However, a neural network was created and trained on 128,175 images and had similar performance to ophthalmologists.\cite{yu2018artificial} The second area AI can help out in clinical practice is in administration. The business of healthcare has become more complex with healthcare infrastructure being stretched thin due to administrative burdens and resourcing constraints. Artificial intelligence can perform repetitive and routine tasks such as patient data entry and automated review of laboratory data and imaging results. If machine learning algorithms are attached to electronic health records, they can assist clinicians and administrators retrieve context related patient data allowing patients to be better treated. Artificial intelligence can also assist in the logistics of hospitals. They can predict hospital stay time from pre-admission data provided by patients, enabling more efficient use of hospital resources. Artificial intelligence can improve the health of patients through patient monitoring. With the adoption of smartphones and fitness monitoring devices, it is now possible to have access to details of patients’ sleep patterns, blood pressure, heart rate, and other measures that was not possible before. These measures when analyzed by medical professionals can tell how healthy or unhealthy someone is, but there is nowhere close to enough people to analyze that much data. AI when trained can analyze this data and extract actionable information.\cite{reddy2019artificial}

		Artificial Intelligence can also be very useful in translational research, the research of turning biomedical research into new therapies, medical procedures, or diagonsistics. Machine learning has been used for biomarker discovery. Biomarkers are measurable indicators of the severity or presence of a disease. Biomarker discovery depends on the identification of unrecognized correlations between thousands of measurements and phenotypes. It is almost impossible for researchers to manually analyze the vast amount of data to find the correlations necessary to define a biomarker. Many of the biomarker panels that have been generated by machine learning have outperformed selected by experts or traditional statistical methods. Artificial intelligence can be used to predict clinical outcomes. Electronic health records with the use of machine learning predict mortality, readmission, and length of hospital stay. If data from health insurance claims are used, the mortality of elderly patients can be predicted.\cite{reddy2019artificial}

		Artificial Intelligence can vastly improve the healthcare provided to patients. Through machine learning, AI has the power to find connections that were not possible before or at least very resource intensive. It can also bring expert level medical opinions to the average person's fingertips. The benefits of AI could take the health of the world to another level.

	\section{Risks of Medical Machine Learning}
		Within public media, perception of the risk of ML is often centered entirely around the threat of some kind of AI uprising.\cite{bbc2016rroai} In contrast, the largest threats that concern professionals are, shown by a survey conducted by Danial Faggella in 2019, the lack of explainability and verifiability in most machine learning models complicated enough to be of use in medicine, along with (though not strictly related to medicine) the potential socio-economic and environmental impact of increased use of machine learning.\cite{emerj2019roawrtiwwa}

@@

		\subsection{Low Explainability in Sufficiently Complex Systems}
			A key issue lies in two important facets of machine learning: explainability of machine learning models and interpretability of their results. Luo \emph{et al.} define explainability as the ability to "summarize the reasons for the behavior of [machine learning] algorithms" and interpretability as the ability to "comprehend what a model did."\cite{doi:10.1259/bjro.20190021} Though the difference between the two is subtle, the distinction is still an important one to make; methods that may increase interpretability may not yield results regarding the overall explainability of the model. However, improvements to explainability increase interpretability by definition.

@@

			\subsubsection{Difficulty of Validation}
				There is immense difficulty of validating the predictions (before acting on them) of a machine learning model. This creates further risk by obfuscating potential error behind a highly technical veil for non-machine learning experts. Moreover, unknown error introduced in input datasets is difficult to detect without access to the data itself, as the ML model will continue to output seemingly valid predictions\cite{10.1145/3328519.3329126} Error could be introduced through any number of innocuous vectors (differences in storage systems between health centers, errors digitizing physical records, decay of electronic records over time) or through more malignant ones (such as a targeted attack). The ethical delicacy of handling electronic health records\cite{ford2016privacy} can make impossible to directly examine new inputs for a source of error, and even if it were possible, the sheer magnitude of the task would be prohibitive. Redyuk \emph{et al.} proposes a method of detecting dataset shifts (without direct examination) that may mitigate this issue in the future\cite{10.1145/3328519.3329126}, but it still remains something important to discuss.

@@

			\subsubsection{Lack of Acceptable Explanation of Decision Making}
				Even if a decision could magically be made with complete and utter certainty at all times, the frequent lack of explainability in machine learning introduces a number of problems and risks at the patient care level. Communication is incredibly important in a doctor-patient relationship,\cite{ha2010doctor} and obfuscating decision making behind a black box harms the ability for clinicians to effectively communicate with patients. When the medical professional does not know how a decision was made, how can they be expected to provide a satisfactory explanation for their patients?

				The need for explanation also highlights an important ethical consideration: what potential aspects of a decision are suitable for an ethical explanation, and on what level should unethical aspects be removed or recontextualized? For instance, suppose skin tone were an emergent feature in a machine learning model; this would clearly be an unacceptable facet of a patient to consider. Would the only proper solution be to accept any potential loss in accuracy and eliminate this feature from the model, or would it be ethical to allow it to remain and simply highlight other features in an explanation given to an inquiring patient? How will the ethical acceptability of given terms affect the architecture of machine learning applications moving forward?

			\subsubsection{Diffusion of Liability}
				Though largely considering more concrete applications of machine learning (such as self driving cars), Reed {et al.}'s 2016 article on the legal responsibility and liability (with regards to English law) points out that there are several parties that could be considered at fault should harm come to a patient due to the complex nature of development and usage of machine learning models. Moreover they posit that it would be incredibly difficult to establish the fault of one party and significantly more expensive to establish the fault of several parties in unison.\cite{reed2016responsibility}

@@

				Methods of responding to these issues are being developed. For instance the GDPR includes a "right to explanation" of algorithmic decisions.\cite{goodman2017european} However, as Watcher \emph{et al.} point out the current state of the GDPR (as of 2017) does not actually ensure a "right to explanation" and instead grants something more akin to a "right to be informed."\cite{wachter2017right} While a true "right to explanation" would be vastly more preferable, it is still a step in the right direction enacted in a very real and actionable way.

@@

		%\subsection{Balance of Regulation Against the Free Market}
			% This section will go into detail on the pros and cons of high and low regulation, emphasizing both the massive increase in risk that comes with low regulations and how high regulations could strangle innovation.

	\section{Conclusion}
		Machine learning is a growing part of health care, and is already seeing significant use within some fields.\cite{Shameer1156} There are many ways we can benefit from advancements in machine learning, however machine learning as a whole is often difficult for non-machine learning experts to understand.\cite{10.1145/2858036.2858529} Improvements in explainability will likely raise a number of ethical conundrums that will significantly affect the shape the architecture of machine learning applications takes. Additionally, while steps are being made to address this on a legal level\cite{goodman2017european} it is clear that we are not quite there yet.\cite{wachter2017right} We are left with an important question: how will machine learning develop within the following years and how will we address the ethical concerns brought forth in the present?
